MWA tied-array processing III: Microsecond time resolution via a synthesis filter

### 2019-10-23 ipfb.tex
Finished(?) changing notation throughout

Like the DFT, the PFB is a transformation from the time domain, $x[n]$, to the frequency domain $X[k]$, where the $[\cdot]$ notation denotes a discretised index, and where $n,k \in \mathbb{Z}$ are the indices for time and frequency, respectively.
Let $N$ be the number of (equally spaced) frequency channels required for some desired spectral resolution.
Although applying a DFT to $N$ adjacent time samples will produce the desired resolution, the result will be an imperfect representation of the true spectral content of $x[n]$ because of \textit{spectral leakage}, which is when power that properly `belongs' to some particular frequency bin appears in (or, is \textit{aliased} to) other, nearby bins.
The impossibility of perfectly eliminating spectral leakage can be seen by realising that operating on a finite-length time series is equivalent to multiplying an arbitrarily long time series with a rectangular window function,
\begin{equation}
    w_R[n] = \begin{cases} 1, & 0 \le n < N, \\ 0, & \text{elsewhere}, \end{cases}
\end{equation}
whose effect in the frequency domain is to convolve the ``true'' spectrum of the signal with the Fourier transform of the window function.
In the case of a rectangular window, this is the sinc function.

A common strategy for mitigating spectral leakage is to choose an alternative window function whose Fourier pair is localised in the frequency domain, and which therefore produces a tolerable level of spectral leakage when convolved with the signal's spectrum.
Many possible windowing functions have been identified, which in general trade the `amount' of leakage with the `location' of the leaked components.
For our purposes, it is sufficient to note that the inevitable presence of a windowing function motivates the definition of the \textit{analysis filter}, $h[n]$, and the generalised \textit{windowed DFT}
\begin{equation}
    X_k[m] = \sum_{n=0}^{K-1} h[n]\,x[n+mM]\,e^{-2\pi jkn/K},
    \label{eqn:wdft}
\end{equation}
where $j = \sqrt{-1}$ denotes the imaginary number, $m$ is the time index of the channelised (output) data, and $0 \le k < K$ denotes the (fine) channel number.
Eq. \eqref{eqn:wdft} describes the action of performing DFTs on short, windowed segments of the input time series of length $K$.
$M$ is the amount that the window is slid along $x[n]$ between successive DFT operations.
If $M < K$ then the windows overlap and the resulting channelisation is \textit{oversampled}; if $M = K$, then it is \textit{critically sampled}.
The choice of $h[n]$ is motivated by the shape of its \textit{frequency response} (i.e. its Fourier pair), whose characteristics (e.g. width, location of sidelobes) might carry particular advantages in certain contexts.
Note that leaving $x[n]$ ``unweighted'' is equivalent to choosing $h[n] = w_R[n]$.

It is well known that scaling a function in the time domain produces the inverse scaling in the Fourier domain.
This fact motivates an alternative strategy for mitigating spectral leakage.
If more samples are available, say $M = NP$ for $P > 1$, then stretching (i.e. scaling and upsampling) the analysis filter will result in a frequency response function that is similar in shape, but $P$ times narrower than the original analysis filter.
Applying the DFT to the larger set of windowed samples and subsequently \emph{downsampling} the resulting spectrum (by choosing every $P$th frequency bin and rejecting the rest) ensures that the original desired spectral resolution is retained, but with significantly less spectral leakage.

The algorithm embodied in the above strategy defines the PFB.
It is clear that choosing a larger value for $P$ will lessen spectral leakage by the same factor, but this naturally comes at a greater computational cost, since the DFT is now applied to $M$ samples.
However, one of the great advantages of the PFB is that the above operations of using a wider analysis wilter and downsampling the resulting spectrum are mathematically equivalent to first segmenting the windowed time series into $P$ segments (known as `taps') of size $N$, summing their respective samples element-wise, and performing a single DFT on the resulting array (now of size $N$), i.e.
\begin{equation}
    X[k] = \sum_{n=0}^{N-1} b[n]\,e^{-2\pi j k (n/N)},
    \label{eqn:pfb}
\end{equation}
where
\begin{equation*}
    b[n] = \sum_{\rho=0}^{P-1} h[n + N\rho]\,x[n + N\rho].
\end{equation*}
A short proof of the equivalence of this expression with the algorithm described above is given in \citet{Harris2011}.
The term `polyphase' derives from the structure of $b[n]$ defined in Eq. \eqref{eqn:pfb}.
It is illustrated in Fig. \ref{fig:pfb}.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{example-image-a}
    \caption{A diagramatic representation of the PFB algorithm, as defined in Eq. \eqref{eqn:pfb}.}
    \label{fig:pfb}
\end{figure}

@@

The first stage (coarse) PFB is described in \citet{Prabu2015}, and here we only document a few details pertaining to the second stage (fine) PFB.
A PFB is specified by (1) the number of output channels, $N$, (2) the number of taps, $P$, and (3) the analysis filter, $h[m]$, $0 \le m < M = NP = 1536$.
For the MWA, $N=128$ (giving fine channels $10\,$kHz wide), $P=12$, and
\begin{equation}
    h[m] = w_H[m]\,w_s[m],
    \label{eqn:filter}
\end{equation}
where
\begin{equation*}
    w_H[m] = \sin^2\bigg(\frac{\pi (m+1)}{M+1}\bigg)
\end{equation*}
is the Hanning window, and
\begin{equation*}
    w_s[m] = \sinc\bigg(\frac{\pi(m + 1 - M/2)}{N}\bigg)
\end{equation*}
is the scaled $\sinc(x) = \sin(x)/x$ function.
It can be easily checked that $h[m]$ is defined to be symmetric around sample $m = 767 = M/2-1$.
The analysis filter is shown in Fig. \ref{fig:filter}.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{filter}
    \caption{Top: The coefficients of the MWA's fine PFB analysis filter, defined in Eq. \eqref{eqn:filter}, which is composed of a Hanning window multiplied to a $\sinc$ function. Bottom: The frequency response of the analysis filter (black, solid), showing negligible attenuation across approximately $10\,$kHz (the bandwidth of a fine channel) and strong attenuation elsewhere. The frequency response is repeated for adjacent channels on either side (grey, dashed) showing crossover points on the channel edges at $-3\,$dB.}
    \label{fig:filter}
\end{figure}

The MWA's fine PFB is \textit{critically sampled}, meaning that consecutive $100\,\mu$s output samples are obtained by applying the analysis filter to blocks of samples separated by the length of one tap, i.e. $N = 128$ samples.
This implies that each coarse channel sample becomes part of the input (i.e. a term in the sum) for $P$ consecutive fine channel samples.
The term ``critically'' sampled is justified by noting that the downsampling by a factor of $P$ (inherent in the PFB algorithm) is exactly counterbalanced by the inclusion of each coarse channel sample in $P$ distinct output channels, with a different weighting (filter coefficient and phase term) applied in each case.
As discussed below, the fact that the PFB is critically sampled informs the design of the synthesis filter used to reconstruct the coarse channel samples from the fine channels.

The rationale behind the particular design choices for the MWA's fine PFB is beyond the scope of this paper---for the purposes of creating a synthesis filter, it is sufficient to know merely how the analysis filter is defined\footnote{Eq. \eqref{eqn:filter} is defined on $\mathbb{R}$, but the actual implementation on the MWA's filed-programmable gate arrays (FPGAs) defines the analysis filter coefficients on $\mathbb{Z}$. The exact implemented values, $h^\ast[m]$, can be obtained from $h[m]$ via $h^\ast[m] \equiv \lfloor\alpha h[m]\rceil$, where $\lfloor\cdot\rceil$ denotes rounding to the nearest integer and $\alpha = 117963.875$.}, and the fact that the PFB is critically sampled.

>>>>

The PFB is a transformation from the time domain, $x[n]$, to the frequency domain $X_k[m]$, where the $[\cdot]$ notation denotes a discretised index, $k$ is the channel number, and $n,m \in \mathbb{Z}$ are the time indices for the pre- and post-channelised data, respectively.
Let $K$ be the number of (equally spaced) frequency channels required for some desired spectral resolution.
Although applying a DFT to $N = K$ adjacent time samples in $x[n]$ will produce the desired resolution, the result will be an imperfect representation of the true spectrum because of \textit{spectral leakage}, which is when power that properly `belongs' to some particular frequency bin appears in (or, is \textit{aliased} to) other, nearby bins.
The impossibility of perfectly eliminating spectral leakage can be seen by realising that operating on a finite-length time series is equivalent to multiplying an arbitrarily long time series with a rectangular window function,
\begin{equation}
    w_R[n] = \begin{cases} 1, & 0 \le n < N, \\ 0, & \text{elsewhere}, \end{cases}
\end{equation}
whose effect in the frequency domain is to convolve the ``true'' spectrum of the signal with the Fourier transform of the window function.
In the case of a rectangular window, this is the sinc function.

A common strategy for mitigating spectral leakage is to choose an alternative window function whose Fourier pair is localised in the frequency domain, and which therefore produces a tolerable level of spectral leakage when convolved with the signal's spectrum.
Many possible windowing functions have been identified, which in general trade the `amount' of leakage with the `location' of the leaked components.
For our purposes, it is sufficient to note that the inevitable presence of a windowing function motivates the definition of the \textit{analysis filter}, $h[n]$, and the generalised \textit{windowed DFT}
\begin{equation}
    X_k[m] = \sum_{n=0}^{N-1} h[n]\,x[n+mM]\,e^{-2\pi jkn/K},
    \label{eqn:wdft}
\end{equation}
where $j = \sqrt{-1}$ denotes the imaginary number, $m$ is the time index of the channelised (output) data, and $0 \le k < K$ denotes the channel number.
Eq. \eqref{eqn:wdft} describes the action of performing DFTs on short, windowed segments of the input time series of length $K$.
$M$ is the amount that the window is slid along $x[n]$ between successive DFT operations.
If $M < K$ then the windows overlap and the resulting channelisation is \textit{oversampled}; if $M = K$, then it is \textit{critically sampled}.
The choice of $h[n]$ is motivated by the shape of its \textit{frequency response} (i.e. its Fourier pair), whose characteristics (e.g. width, location of sidelobes) are chosen according to the advantages they carry in particular contexts.
Leaving $x[n]$ ``unweighted'' is equivalent to choosing $h[n] = w_R[n]$, in which case Eq. \eqref{eqn:wdft} is also called the short-time Fourier transform (STFT).
On the other hand, choosing $h[n]$ to be the $\sinc$ function will result in a frequency response that approximates a rectangular window.

It is well known that scaling a function in the time domain produces the inverse scaling in the Fourier domain.
This fact motivates an alternative strategy for mitigating spectral leakage.
Choosing a larger window size, $N = KP$ (for integer $P > 1$), and a corresponding wider analysis filter, will result in a frequency response function that is similar in shape, but $P$ times narrower than the original analysis filter.
A DFT applied to the larger number of samples will naturally produce a correspondingly larger number of frequency channels, but choosing only every $P$th channel and discarding the rest (known as \textit{decimation}) ensures that the desired spectral resolution with $K$ channels is retained.
In this way, spectral leakage can be contained arbitrarily close to the ``correct'' channel by choosing a sufficiently high value of $P$.

The two-step algorithm described above (performing a windowed DFT on $N = KP$ samples and decimating the resulting spectrum) defines the PFB.
Formally, it is equivalent to Eq. \eqref{eqn:wdft}; however, the term \textit{critically sampled} (i.e. $M = K$) implies that the $N$-length windows will now overlap.
The term ``polyphase'' derives from the fact that each block of $K = N/P$ samples (knows as \textit{taps}) in $x[n]$ is included in multiple applications of the DFT, but appearing at a different relative phase in each case.

One of the great advantages of the PFB is the existence of a mathematically equivalent but computationally efficient implementation.
It can be shown that Eq. \eqref{eqn:wdft} is equivalent to first segmenting the windowed time series into taps, summing their respective samples element-wise, and performing a single DFT on the resulting array (now also of size $K$), i.e.
\begin{equation}
    X_k[m] = \sum_{n=0}^{K-1} b[n]\,e^{-2\pi jkn/K},
    \label{eqn:pfb}
\end{equation}
where
\begin{equation*}
    b[n] = \sum_{\rho=0}^{P-1} h[n + K\rho]\,x[n + mM + K\rho].
\end{equation*}
A short proof of this is given in \citet{Harris2011}.
Eq. \eqref{eqn:pfb} is illustrated in Fig. \ref{fig:pfb}.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{example-image-a}
    \caption{A diagramatic representation of the PFB algorithm, as defined in Eq. \eqref{eqn:pfb}.}
    \label{fig:pfb}
\end{figure}

@@

The first stage (coarse) PFB is described in \citet{Prabu2015}, and here we only document a few details pertaining to the second stage (fine) PFB.
A PFB is specified by (1) the number of output channels, $K$, (2) the number of taps, $P$, and (3) the analysis filter, $h[n]$.
For the MWA, $K = 128$ (giving fine channels $10\,$kHz wide), $P = 12$, and
\begin{equation}
    h[n] = w_H[n]\,w_s[n],
    \label{eqn:filter}
\end{equation}
where
\begin{equation*}
    w_H[n] = \sin^2\bigg(\frac{\pi (n+1)}{N+1}\bigg)
\end{equation*}
is the Hanning window, and
\begin{equation*}
    w_s[n] = \sinc\bigg(\frac{\pi(n + 1 - N/2)}{K}\bigg)
\end{equation*}
is the scaled $\sinc(x) = \sin(x)/x$ function.
It can be easily checked that $h[m]$ is defined to be symmetric around sample $m = 767 = M/2-1$.
The analysis filter is shown in Fig. \ref{fig:filter}.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{filter}
    \caption{Top: The coefficients of the MWA's fine PFB analysis filter, defined in Eq. \eqref{eqn:filter}, which is composed of a Hanning window multiplied to a $\sinc$ function. Bottom: The frequency response of the analysis filter (black, solid), showing negligible attenuation across approximately $10\,$kHz (the bandwidth of a fine channel) and strong attenuation elsewhere. The frequency response is repeated for adjacent channels on either side (grey, dashed) showing crossover points on the channel edges at $-3\,$dB.}
    \label{fig:filter}
\end{figure}
The MWA's fine PFB is critically sampled, with $M = K = 128$.
The rationale behind the particular design choices for the MWA's fine PFB is beyond the scope of this paper---for the purposes of creating a synthesis filter, it is sufficient to know merely how the analysis filter is defined\footnote{Eq. \eqref{eqn:filter} is defined on $\mathbb{R}$, but the actual implementation on the MWA's filed-programmable gate arrays (FPGAs) defines the analysis filter coefficients on $\mathbb{Z}$. The exact implemented values, $h^\ast[n]$, can be obtained from $h[n]$ via $h^\ast[n] \equiv \lfloor\alpha h[n]\rceil$, where $\lfloor\cdot\rceil$ denotes rounding to the nearest integer and $\alpha = 117963.875$.}, and the fact that the PFB is critically sampled.
